---
title: "Introduction"
author: "Shian Su"
output: html_document
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# preload to avoid loading messages
library(NanoMethViz)
```

```{r}
library(NanoMethViz)
```

To generate a methylation plot we need 3 components:

* methylation data in tabix format
* annotation of exons
* annotation of samples

The methylation information has been modified from the output of nanopolish/f5c. It has then been compressed and indexed using `bgzip()` and `indexTabix()` from the `Rsamtools` package.

```{r}
# methylation data stored in tabix file
methy <- system.file(package = "NanoMethViz", "methy_subset.tsv.bgz")

# tabix is just a special gzipped tab-separated-values file
read.table(gzfile(methy), col.names = methy_col_names(), nrows = 6)
```

The exon annotation was obtained from the Mus.musculus package, and joined into a single table. It is important that the chromosomes share the same convention as that found in the methylation data.

```{r}
# helper function extracts exons from Mus.muculus pacakge
exon_tibble <- get_exons_mus_musculus()

head(exon_tibble)
```

We will defined the sample annotation ourselves. It is important that the sample names match those found in the methylation data.

```{r}
sample <- c(
  "B6Cast_Prom_1_bl6",
  "B6Cast_Prom_1_cast",
  "B6Cast_Prom_2_bl6",
  "B6Cast_Prom_2_cast",
  "B6Cast_Prom_3_bl6",
  "B6Cast_Prom_3_cast"
)

group <- c(
  "bl6",
  "cast",
  "bl6",
  "cast",
  "bl6",
  "cast"
)

sample_anno <- data.frame(sample, group, stringsAsFactors = FALSE)

sample_anno
```

For convenience we assemble these three pieces of data into a single object.

```{r}
nmeth_results <- NanoMethResult(methy, sample_anno, exon_tibble)
```

The genes we have available are

* Peg3
* Meg3
* Impact
* Xist
* Brca1
* Brca2

For demonstrative purposes we will plot Peg3.

```{r}
plot_gene(nmeth_results, "Peg3")
```

We can also load in some DMR results to highlight DMR regions.

```{r}
# loading saved results from previous bsseq analysis
bsseq_dmr <- read.table(
    system.file(package = "NanoMethViz", "dmr_subset.tsv.gz"),
    sep = "\t",
    header = TRUE,
    stringsAsFactors = FALSE
)
```

```{r}
plot_gene(nmeth_results, "Peg3", anno_regions = bsseq_dmr)
```

Individual long reads can be visualised using the `spaghetti` argument.

```{r, warning = FALSE}
# warnings have been turned off in this vignette, but this will generally
# generate many warnings as the smoothing for many reads will fail
plot_gene(nmeth_results, "Peg3", anno_regions = bsseq_dmr, spaghetti = TRUE)
```

